Computing volume via oblique or irregularly shaped rectangular infinitesimals

### 2019-08-19 Cav2D/article-template.tex
added some references

\title{Computing volume via non-uniform rectangular infinitesimals}
%\author{Woodrow Wilson and Herbert Hoover} %%leave blank in initial submission to allow for double blind reviewing
%\author{ }

@@

%\item[\biogpic{\includegraphics[width=84pt]{Herbert_Hoover.pdf}}Herbert Hoover] (hchoover@stanford.edu) entered Stanford University in 1891, after failing all of the entrance exams except mathematics.  He received his BS degree in geology in 1895, spent time as a mining engineer, then was appointed by his co-author to the U.S. Food Administration and the Supreme Economic Council, where he orchestrated the greatest famine relief efforts of all time.
%\end{biog}
%This is not to say that there  This is partly, because there is no simplistic geometric interpretation for fractional calculus.  
\noindent
Bonaventura Cavalieri was a mathematician who lived in the 17th century. He made numerous contributions to the fields of optics, astronomy and geometry.
He served as a professor of mathematics at the university of Bologna from 1629 until his death in 1647. 
Cavalieri was a student of Galileo. Galileo held Cavalieri in high esteem. Galileo made the following remark in a reference letter he wrote on Cavalieri's behalf: ``\emph{few, perhaps none, from 
Archimedes onwards, have investigated so deeply and have entered so far into the science of geometry}''. In 1935, a moon crater, Cavalerius, was officially named after Cavalieri to commemorate his achievements and to honor his memory (see Figure~\ref{fig:crater}) \cite{?}. 
\begin{figure}[htb]
\centering
\includegraphics[width=0.4\textwidth]{crater.jpg}
\caption{The moon crater Cavalerius, which is named after Bonaventura Cavalieri.}
\label{fig:crater}
\end{figure}
Cavalieri's most notable achievement was the publication of his 
treatise, \emph{Geometria indivisibilibus}, in which he describes the precalculus method of indivisibles. The method of indivisibles can be attributed to the ancient Greeks,
in this case Democritus and Archimedes. The method is used to compute the area and the volume of complex planar figures and solids, respectively.\\

\noindent
Two planar figures or solids are Cavalieri congruent if they can be decomposed into the same set of parallel ``indivisibles''. An indivisible of 
a planar figure is any cord of the planar figure. An indivisible of a solid is any planar section of a solid. Two indivisibles are equal if they are of the same length or area.
It is Cavalieri's principle that assures us that two planar figures or solids that are composed of the same set of indivisibles have equal area or volume. Cavalieri's principle 
in $\mathbb{R}^3$ is stated below without proof:

\begin{theorem}[Cavalieri's principle in $\mathbb{R}^3$]
If two solids are included between a pair of parallel planes, and if the areas of
the two sections cut by them on any plane parallel to the including planes are always
equal, then the volumes of the two solids are equal.
\end{theorem}
\noindent
The method of indivisibles works by finding a planar figure or a solid, whose area or volume can easily be computed, which is Cavalieri congruent to a planar figure or solid whose 
area or volume is difficult to compute.

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{cav_principle.pdf}
\caption{Illustration of Cavalier's principle in $\mathbb{R}^3$. The volume of the left rectangular prism is equal to the volume of the right non-uniform reectangular prism 
due to Cavalieri's principle in $\mathbb{R}^3$. The infinitesimals associated with the two solids are also depicted.}
\label{fig:cav_principle}
\end{figure}

\noindent
Figure~\ref{fig:cav_principle} depicts a classic rectangular prism or column (left solid) as well as a \emph{non-uniform rectangular prism or column} (right solid). The volumes of the two solids in Figure~\ref{fig:cav_principle} are equal due 
to Cavalieri's principle. Note that in this article we use the non-standardized phrase non-unifor rectangular prism or column to refer to a solid that has rectangular 
basis and lateral edges that are either oblique lines or non-linear curves. Moreover, the four lateral edges of the aformentioned solid also need to be translations of one another. In this paper, we will also take the liberty to refer to a non-uniform rectangular prism whose base's
lenght and widht both tend to zero as a \emph{non-uniform rectangular infinitesimal}. \\

\noindent
Ackermann et al. utilized the idea of using non-rectangular integration strips to perform integration in $\mathbb{R}^2$ and in doing so developed the Cavalieri integral \cite{ackermann12}. 
The area of a non-rectangular integration strip is computed via Cavalieri’s principle. The Cavalieri integral has proven to be very useful: it can be used to visualize the Riemann-Stieltjes integral \cite{grobler19}. In this paper, we extend the Cavalieri integral into $\mathbb{R}^3$, i.e. we use Cavalieri's principle in $\mathbb{R}^3$ to define the double 
Cavalieri integral. We also show that if a solid can be decomposed into an infinite number of non-uniform rectangular infinitesimals then its volume can be computed 
by the double Cavalieri integral. Whether the double Cavalieri integral will prove to be an equally useful visualization tool will be investigated as part of a future endeavor.
Moreover, the examples we provide at the end of the paper are presented in such a way that they will deepen undergraduate Calculus students' understanding into the 
process of integration. Furhtermore, we will try to steer clear from using too many formal definitions and proofs and rather convey our ideas via worked out examples, as this paper is mainly meant to be read by undergraduate students as well as the layman who is interested in the theory of integration.\\

@@

\section{Cavalieri Integral}
The single Cavalieri integral was originally defined by Ackermann et al. without having to resort to vector calculus \cite{ackermann12}. In this section, however, we will define the Cavalieri integral using vector calculus notation. As the reader will undoubtedly realize, if we define the single Cavalieri integral using vector calculus then it becomes trivial to
extend its definition into $\mathbb{R}^3$.\\

@@

\section{Double Cavalieri Integral}
In this section we define the double Cavalieri integral and we show that it can be used to calculate the volume of solids that can be decomposed into an infinite 
number of non-uniform regtangular infinitesimals.\\

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. Note that 
$\mathcal{S}$ can be decomposed into an infinite number of non-uniform rectangular infinitesimals. As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
In general, let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\ref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_j^1$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_j^1$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_j^2)$ represents the height of a non-uniform rectangular column or prism $\mathcal{K}_{ij}$ whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_j^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\section{Prismatoid}
Let $\mathcal{S}$ be the solid which is bounded by the planes $f(x,y)=-x-y+8$, $2x$, $2(x-4)$, $y$, $y-4$ and the $xy$-plane. Solid $\mathcal{S}$ is depicted 
in Figure~\ref{fig:prismatoid_solid}, Figure~\ref{fig:prismatoid} and Figure~\ref{fig:prismatoid_real}. Solid $\mathcal{S}$ is a prismatoid, since all of its 
vertices lie in two parallel planes. Furthermore, all of its lateral faces are trapezoids or triangles (if viewed from the top). Also note the following: the base of $\mathcal{S}$ is the region 
 $R = [0,4]\times [0,4]$ (if viewed from the front) and that the line of intersection between $2x$ and $y$ is $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$. The region $R$ and the line $\mathbf{c}(t)$ are
also depicted and labelled in Figure~\ref{fig:prismatoid_solid}.\\

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{prismatoid.pdf}
\caption{Another depiction of the solid $\mathcal{S}$ found in Figure~\ref{fig:prismatoid_solid} which highlights the fact that $\mathcal{S}$ is a prismatoid. We can, therefore, calculate its volume with the prismatoid volume equation: $V_{\mathcal{S}}= \frac{h}{6}(A_1+4A_2+A_3)$.
Note, the area of the two basis of $\mathcal{S}$ are denoted by $A1$ and $A3$, the cross-sectional area of the intersection of $\mathcal{S}$ with a plane midway between the two basis is denoted via $A2$, 
and the height of $\mathcal{S}$ is denoted by $h$. Moreover, also note that the surfaces associated with $A1$, $A2$ and $A3$ are depicted in the above figure using shaded sheets. For $\mathcal{S}$ 
$A1=10.7331$, $A2= 7.1554$, $A3 = 3.5777$ and $h = 3.5777$. The volume of $\mathcal{S}$ is, therefore, $25.6$.}
\label{fig:prismatoid}
\end{figure}

\noindent
As we alluded to in the previous section we can calculate $V_{\mathcal{S}}$ using the double Cavalieri integral:
\begin{equation}
\label{eq:cav_ex1}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We can, however, only evaluate the above integral if the transformation $T$ is known (see Figure~\ref{fig:prismatoid_regions}). We can find $T$ using Equation~\eqref{eq:h_2d} and 
Equation~\eqref{eq:s_2d}. To evaluate Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} we first need to express $t$ in terms of $x_i^1$ and $y_j^1$. If we substitute $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t(x_i^1,y_j^1) = \frac{x_i^1}{5} + \frac{y_j^1}{5} + \frac{8}{5}. 
\end{equation}
Substituting $t(x_i^1,y_j^1)$ into Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} results in:
\begin{equation}
x_{i}^2 = h(x_{i}^1,y_{j}^1) = \frac{4}{5} x_{i}^1 - \frac{1}{5} y_{j}^1 + \frac{8}{5}  
\end{equation}
and
\begin{equation}
y_{j}^2 = r(x_{i}^1,y_{j}^1) = \frac{3}{5} x_{i}^1 - \frac{2}{5} y_{j}^1 + \frac{16}{5}.  
\end{equation}
We can now evaluate Equation~\eqref{eq:cav_ex1}:
\begin{eqnarray}
\label{eq:final_int_prismatoid}
V_{\mathcal{S}}=\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^4\int_0^4 f(h(x,y),r(x,y))~dydx\\
&=& \int_0^4 \int_0^4 -\frac{7}{5}x + \frac{3}{5} y + \frac{16}{5}~dydx\\
&=& 25.6
\end{eqnarray}

\noindent
We can easily verify that the $V_{\mathcal{S}} = 25.6$ is in fact correct, since $\mathcal{S}$ is a prismatoid. According to the prismatoid volume 
equation $V_{\mathcal{S}}=25.6$. The details of this computation is described in the caption of Figure~\ref{fig:prismatoid}. This then validates Equation~\ref{eq:final_int_prismatoid}.

@@

\noindent
Since $\mathcal{S}$ can be decomposed into an infinite number of non-uniform rectangular infinitesimals we can compute its volume using:
\begin{equation}
\label{eq:cav_ex2}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We, however, first need to calculate $T$ if we wish to evaluate Equation~\ref{eq:cav_ex2} (see Figure~\ref{fig:sphere_regions}). We can determine $T$ via 
Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d}. Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} in turn requires $t(x_i^1,y_j^2)$, which can be obtained via Equation~\eqref{eq:c_z}.
If we substite $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t = \frac{1}{4}[-2 x_{i}^1 - 2 y_{j}^1  -  1 + K(x_{i}^1,y_{j}^1)],
\end{equation}
where
\begin{equation}
K(x_{i}^1,y_{j}^1) = \sqrt{-4(x_{i}^1)^2+4x_{i}^1+8x_{i}^1y_{j}^1-4(y_{j}^1)^2+4y_{j}^1 + 33}. 
\end{equation}
As mentioned earlier, we can use $t(x_i^1,y_j^1)$ to find expressions for $h$ and $s$. The expressions of $h$ and $s$ so obtained can then be used to evaluate Equation~\eqref{eq:cav_ex2}:  
\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^1\int_0^1 f(h(x,y),r(x,y))~dydx\\
&=& \int_0^1 \int_0^1 \sqrt{-(x+t(x,y))^2 - (y+t(x,y))^2 + 4} ~dydx\\
&=& 0.864043
\end{eqnarray}

@@

\caption{Computing the volume of the solid $\mathcal{S}$ depicted in Figure~\ref{fig:sphere_solid} using a more traditional approach. The region $R_k$ is bounded by $x=a$, $x=b$, $g_1(x)$ and $g_2(x)$. Moreover, $c_1 \approx 1.1861407$, $c_2 \approx 1.6861407$, $c_3 \approx 0.6861407$ and $c_4 \approx 1.3507811$. $r_1(x) = \frac{1}{2}[\sqrt{17-4x^2}-1]$,
$r_2(x) = \sqrt{-x^2-x+4}$, $r_3(x) = \sqrt{-x^2-(x-1)+4}$ and $r_4(x) = \frac{1}{2}[\sqrt{21-4x^2}-1]$. If we sum the last column we obtain the volume of $\mathcal{S}$ which is $0.864043$.} 
\label{tab:table1}
\begin{center}

@@

\begin{abstract}
In this paper we extended the Cavalieri integral into three dimensional euclidean space, i.e. we defined the double Cavalieri integral. 
We also showed that this integral can be used to compute the volume of solids that can be decomposed into an infinite number of 
non-uniform rectangular infinitesimals. We end the paper with two fully worked out examples. In the first example we compute the 
volume of a prismatoid using the double Cavalieri integral. In the second example we compute the volume of a complex solid which is housed within a sphere using the double Cavalieri integral.
\end{abstract}

@@

\bibitem{adda97} Adda,~F.~B. (1997). Geometric interpretation of the fractional derivative. \textit{J. of Fract. Calc.}, 11: 21--52.

\bibitem{bartle76} Bartle,~R.~G. (1976). \textit{The Elements of Real Analysis.} New York: Wiley.

%Physical
\bibitem{cioc16} Cioc, R. (2016). Physical and geometrical interpretation of Gr\"{u}nwald-Letnikov differintegrals: Measurement of path and acceleration. \textit{Fract. Calc. Appl. Anal.} 19(1): 161--172.

\bibitem{davis59} Davis,~P.~J. (1959). Leonhard Euler's Integral: A Historical Profile of the Gamma Function. \textit{Amer. Math. Monthly}. 66(10): 849--869

\bibitem{euler1738} Euler,~L. (1738). De progressionibus transcendentibus seu quarum termini generales algebraice dari nequeunt. \textit{Commentarii academiae scientiarum Petropolitanae}. 36--57.

\bibitem{gomez14} Gomez-Aguilara,~J.~F., Razo-Hernandez,~R., Granados-Lieberman, D.~ A (2014). Physical interpretation of fractional
calculus in observable terms: analysis of the fractional time constant and the transitory response. \textit{Revista Mexicana de Fisica.} 60: 32--38

>>>>

\title{Computing volume via oblique or irregularly shaped rectangular infinitesimals}
%\author{Woodrow Wilson and Herbert Hoover} %%leave blank in initial submission to allow for double blind reviewing
%\author{ }

@@

%\item[\biogpic{\includegraphics[width=84pt]{Herbert_Hoover.pdf}}Herbert Hoover] (hchoover@stanford.edu) entered Stanford University in 1891, after failing all of the entrance exams except mathematics.  He received his BS degree in geology in 1895, spent time as a mining engineer, then was appointed by his co-author to the U.S. Food Administration and the Supreme Economic Council, where he orchestrated the greatest famine relief efforts of all time.
%\end{biog}
%This is not to say that there  This is partly, because there is no simplistic geometric interpretation for fractional calculus.  
\noindent
Bonaventura Cavalieri was a mathematician who lived in the 17th century. He made numerous contributions to the fields of optics, astronomy and geometry.
He served as a professor of mathematics at the university of Bologna from 1629 until his death in 1647 \cite{andersen1985,eves1991,jullien2015}. 
Cavalieri was a student of Galileo. Galileo held Cavalieri in high esteem. Galileo made the following remark in a reference letter he wrote on Cavalieri's behalf: ``\emph{few, perhaps none, from 
Archimedes onwards, have investigated so deeply and have entered so far into the science of geometry}'' \cite{gonzalez2018}. In 1935, a moon crater, Cavalerius, was officially named after Cavalieri to commemorate his achievements and to honor his memory (see Figure~\ref{fig:crater}) \cite{blagg1935}. 
\begin{figure}[htb]
\centering
\includegraphics[width=0.4\textwidth]{crater.jpg}
\caption{The moon crater Cavalerius, which is named after Bonaventura Cavalieri. James Stuby / Wikimedia Commons / Public Domain.}
\label{fig:crater}
\end{figure}
Cavalieri's most notable achievement was the publication of his 
treatise, \emph{Geometria indivisibilibus}, in which he describes the precalculus method of indivisibles. The method of indivisibles can be attributed to the ancient Greeks,
in this case Democritus and Archimedes \cite{eves1991}. The method is used to compute the area and the volume of complex planar figures and solids, respectively.\\

\noindent
Two planar figures or solids are Cavalieri congruent if they can be decomposed into the same set of parallel ``indivisibles''. An indivisible of 
a planar figure is any cord of the planar figure. An indivisible of a solid is any planar section of a solid. Two indivisibles are equal if they are of the same length or area.
It is Cavalieri's principle that assures us that two planar figures or solids that are composed of the same set of indivisibles have equal area or volume \cite{andersen1985,eves1991,jullien2015,malik1984,wildberger2002,william1917,young1985}. Cavalieri's principle 
in $\mathbb{R}^3$ is stated below without proof:

\begin{theorem}[Cavalieri's principle in $\mathbb{R}^3$]
If two solids are included between a pair of parallel planes, and if the areas of
the two sections cut by them on any plane parallel to the including planes are always
equal, then the volumes of the two solids are equal (see Figure~\ref{fig:cav_principle}).
\end{theorem}
\noindent
The method of indivisibles works by finding a planar figure or a solid, whose area or volume can easily be computed, which is Cavalieri congruent to a planar figure or solid whose 
area or volume is difficult to compute.

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{cav_principle.pdf}
\caption{Illustration of Cavalier's principle in $\mathbb{R}^3$. The volume of the left rectangular prism is equal to the volume of the right irregularly shaped rectangular prism 
due to Cavalieri's principle in $\mathbb{R}^3$. The infinitesimals associated with these two solids are also depicted.}
\label{fig:cav_principle}
\end{figure}

\noindent
Figure~\ref{fig:cav_principle} depicts a classic rectangular prism or column (left solid) as well as an \emph{irregular rectangular prism or column} (right solid). The volumes of the two solids in Figure~\ref{fig:cav_principle} are equal due 
to Cavalieri's principle. Note that in this article we use the non-standardized phrase irregular rectangular prism or column to refer to a solid that has rectangular 
basis and lateral edges that are non-linear curves. Moreover, the four lateral edges of the aformentioned solid also need to be translations of one another. In this paper, we will also take the liberty to refer to an irregular rectangular prism whose base's
lenght and widht both tend to zero as an \emph{irregular rectangular infinitesimal}. \\

\noindent
Ackermann et al. utilized the idea of using non-rectangular integration strips to perform integration in $\mathbb{R}^2$ and in doing so developed the Cavalieri integral \cite{ackermann12}. 
The area of a non-rectangular integration strip is computed via Cavalieri’s principle. The Cavalieri integral has proven to be very useful: it can be used to visualize the Riemann-Stieltjes integral \cite{grobler19}. In this paper, we extend the Cavalieri integral into $\mathbb{R}^3$, i.e. we use Cavalieri's principle in $\mathbb{R}^3$ to define the double 
Cavalieri integral. We also show that if a solid can be decomposed into an infinite number of irregularly shaped rectangular infinitesimals then its volume can be computed 
by the double Cavalieri integral. Whether the double Cavalieri integral will prove to be an equally useful visualization tool will be investigated as part of a future endeavor.
Moreover, the examples we provide at the end of the paper are presented in such a way that they will deepen undergraduate Calculus students' understanding into the 
process of integration. Furhtermore, as this paper is mainly meant to be read by undergraduate students as well as the layman who is interested in the theory of integration, we will as far as possible try to steer clear from using too many formal definitions and proofs and rather convey our ideas via worked out examples.\\

@@

\section{Cavalieri Integral}
The single Cavalieri integral was originally defined by Ackermann et al. without having to resort to vector calculus \cite{ackermann12,stewart1999}. In this section, however, we will define the Cavalieri integral using vector calculus notation. As the reader will undoubtedly realize, if we define the single Cavalieri integral using vector calculus then it becomes trivial to
extend its definition into $\mathbb{R}^3$.\\

@@

\section{Double Cavalieri Integral}
In this section we define the double Cavalieri integral and we show that it can be used to calculate the volume of solids that can be decomposed into an infinite 
number of irregularly shaped regtangular infinitesimals.\\

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. Note that 
$\mathcal{S}$ can be decomposed into an infinite number of irregularly shaped rectangular infinitesimals. As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
In general, let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\ref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_j^1$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_j^1$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_j^2)$ represents the height of an irregular rectangular column or prism $\mathcal{K}_{ij}$ (or in the case of Figure~\ref{fig:prismatoid_solid} and oblique rectangular prism) whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_j^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\section{Prismatoid}
Let $\mathcal{S}$ be the solid which is bounded by the planes $f(x,y)=-x-y+8$, $2x$, $2(x-4)$, $y$, $y-4$ and the $xy$-plane. Solid $\mathcal{S}$ is depicted 
in Figure~\ref{fig:prismatoid_solid}, Figure~\ref{fig:prismatoid} and Figure~\ref{fig:prismatoid_real}. Solid $\mathcal{S}$ is a prismatoid, since all of its 
vertices lie in two parallel planes. Furthermore, all of its lateral faces are trapezoids or triangles (if viewed from the top) \cite{bradley1979}. Also note the following: the base of $\mathcal{S}$ is the region 
 $R = [0,4]\times [0,4]$ (if viewed from the front) and that the line of intersection between $2x$ and $y$ is $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$. The region $R$ and the line $\mathbf{c}(t)$ are
also depicted and labelled in Figure~\ref{fig:prismatoid_solid}.\\

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{prismatoid.pdf}
\caption{Another depiction of the solid $\mathcal{S}$ found in Figure~\ref{fig:prismatoid_solid} which highlights the fact that $\mathcal{S}$ is a prismatoid. We can, therefore, calculate its volume with the prismatoid volume equation: $V_{\mathcal{S}}= \frac{h}{6}(A_1+4A_2+A_3)$.
Note, the area of the two basis of $\mathcal{S}$ are denoted by $A1$ and $A3$, the cross-sectional area of the intersection of $\mathcal{S}$ with a plane midway between the two basis is denoted via $A2$, 
and the height of $\mathcal{S}$ is denoted by $h$. Moreover, also note that the surfaces associated with $A1$, $A2$ and $A3$ are depicted in the above figure using shaded sheets. For $\mathcal{S}$, 
$A1=10.7331$, $A2= 7.1554$, $A3 = 3.5777$ and $h = 3.5777$. The volume of $\mathcal{S}$ is, therefore, $25.6$.}
\label{fig:prismatoid}
\end{figure}

\noindent
As we alluded to in the previous section we can calculate $V_{\mathcal{S}}$ using the double Cavalieri integral:
\begin{equation}
\label{eq:cav_ex1}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We can, however, only evaluate the above integral if the transformation $T$ is known (see Figure~\ref{fig:prismatoid_regions}). We can find $T$ using Equation~\eqref{eq:h_2d} and 
Equation~\eqref{eq:s_2d}. To evaluate Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} we first need to express $t$ in terms of $x_i^1$ and $y_j^1$. If we substitute $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t(x_i^1,y_j^1) = \frac{1}{5}x_i^1 + \frac{1}{5}y_j^1 + \frac{8}{5}. 
\end{equation}
Substituting $t(x_i^1,y_j^1)$ into Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} results in:
\begin{equation}
x_{i}^2 = h(x_{i}^1,y_{j}^1) = \frac{4}{5} x_{i}^1 - \frac{1}{5} y_{j}^1 + \frac{8}{5}  
\end{equation}
and
\begin{equation}
y_{j}^2 = r(x_{i}^1,y_{j}^1) = \frac{3}{5} x_{i}^1 - \frac{2}{5} y_{j}^1 + \frac{16}{5}.  
\end{equation}
We can now evaluate Equation~\eqref{eq:cav_ex1}:
\begin{eqnarray}
\label{eq:final_int_prismatoid}
V_{\mathcal{S}}=\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^4\int_0^4 f(h(x,y),r(x,y))~dydx\\
&=& \int_0^4 \int_0^4 -\frac{7}{5}x + \frac{3}{5} y + \frac{16}{5}~dydx\\
&=& 25.6
\end{eqnarray}

\noindent
We can easily verify that the $V_{\mathcal{S}} = 25.6$ is in fact correct, since $\mathcal{S}$ is a prismatoid. Indeed, according to the prismatoid volume 
equation,  $V_{\mathcal{S}}=25.6$. The details of this computation is described in the caption of Figure~\ref{fig:prismatoid}.

@@

\noindent
Since $\mathcal{S}$ can be decomposed into an infinite number of irregularly shaped rectangular infinitesimals we can compute its volume using:
\begin{equation}
\label{eq:cav_ex2}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We, however, first need to calculate $T$ if we wish to evaluate Equation~\ref{eq:cav_ex2} (see Figure~\ref{fig:sphere_regions}). We can determine $T$ via 
Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d}. Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} in turn requires $t(x_i^1,y_j^2)$, which can be obtained via Equation~\eqref{eq:c_z}.
If we substite $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t = \frac{1}{4}[-2 x_{i}^1 - 2 y_{j}^1  -  1 + K(x_{i}^1,y_{j}^1)],
\end{equation}
where
\begin{equation}
K(x_{i}^1,y_{j}^1) = \sqrt{-4(x_{i}^1)^2+4x_{i}^1+8x_{i}^1y_{j}^1-4(y_{j}^1)^2+4y_{j}^1 + 33}. 
\end{equation}
As mentioned earlier, we can use $t(x_i^1,y_j^1)$ to find expressions for $h$ and $s$. The expressions of $h$ and $s$ so obtained can then be used to evaluate Equation~\eqref{eq:cav_ex2}:  
\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^1\int_0^1 f(h(x,y),r(x,y))~dydx\\
&=& \int_0^1 \int_0^1 \sqrt{-(x+t(x,y))^2 - (y+t(x,y))^2 + 4} ~dydx\\
&=& 0.864043
\end{eqnarray}

@@

\caption{Computing the volume of the solid $\mathcal{S}$ depicted in Figure~\ref{fig:sphere_solid} using a more traditional approach. The region $R_k$ is bounded by $x=a$, $x=b$, $g_1(x)$ and $g_2(x)$ (see Figure~\ref{fig:sphere_xy}). Moreover, $c_1 \approx 1.1861407$, $c_2 \approx 1.6861407$, $c_3 \approx 0.6861407$ and $c_4 \approx 1.3507811$. $r_1(x) = \frac{1}{2}[\sqrt{17-4x^2}-1]$,
$r_2(x) = \sqrt{-x^2-x+4}$, $r_3(x) = \sqrt{-x^2-(x-1)+4}$ and $r_4(x) = \frac{1}{2}[\sqrt{21-4x^2}-1]$. If we sum the last column we obtain the volume of $\mathcal{S}$ which is $0.864043$.} 
\label{tab:table1}
\begin{center}

@@

\begin{abstract}
In this paper we extended the Cavalieri integral into three dimensional euclidean space, i.e. we defined the double Cavalieri integral. 
We also showed that this integral can be used to compute the volume of solids that can be decomposed into an infinite number of 
irrgularly shaped rectangular infinitesimals. We end the paper with two fully worked out examples. In the first example we compute the 
volume of a prismatoid using the double Cavalieri integral. In the second example we compute the volume of a complex irregular shaped solid which is housed within a sphere using the double Cavalieri integral.
\end{abstract}

@@

\bibitem{andersen1985}
K.~Andersen. (1985). Cavalieri's Method of Indivisibles. \emph{Arch. Hist. Exact Sci.} 31(4): 291--367.

\bibitem{blagg1935}
M.~A Blagg, A. Blagg and K. M{\"u}ller. (1935). \emph{Named Lunar Formations}. Percy Lund, Humphries and Co. Ltd.

\bibitem{bradley1979}
A.~D.~Bradley. (1979). Prismatoid, Prismoid, Generalized Prismoid. \emph{Amer. Math. Monthly.} 86(6): 486--490.

\bibitem{eves1991}
H.~Eves. (1991). Two Surprising Theorems on Cavalieri Congruence. \emph{College Math. J.} 22(2): 118--124.

\bibitem{gonzalez2018} D.~L.~Gonz{\'a}lez, D.~L.~Rabiti, J.~H.~E~Cartwright. 2018. Bonaventura Cavalieri and Bologna. \emph{Math. Intelligencer}. 40(2): 21--29.
